---
title: "Appendix: Figure A1"
subtitle: "Making Predictions When Causal Models Change – The Role of Confounding"
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

This script produces Figure A1 of the manuscript: normative values of P(e|c)
under a generative (Panel A) and a preventive (Panel B) confounding link
A → C, as a function of the base rate of the target cause (b_c) and the
strength of the confounding link (w_ac). Figure output: `figures/`.

# Setup

```{r setup}
library(ggplot2)
library(ggpubr)  # ggarrange()

dir.create("figures", showWarnings = FALSE)
```

# Normative functions

```{r functions}
# Equation 4: generative confounding link A -> C
P_ec_cccf <- function(b_c, w_ac, w_ae, w_ce, b_a) {
  numerator   <- w_ac * b_a + b_c * b_a - w_ac * b_c * b_a
  denominator <- w_ac * b_a + b_c - w_ac * b_c * b_a
  P_ac <- numerator / denominator
  w_ce + (1 - w_ce) * w_ae * P_ac
}

# Equations A1 and A2: preventive confounding link A -> C
# P(a|c) under noisy-AND-NOT; b_c cancels out
P_ec_cc_minus_cf <- function(b_c, w_ac, w_ae, w_ce, b_a) {
  P_ac <- (b_a * (1 - w_ac)) / (1 - b_a * w_ac)
  w_ce + (P_ac * w_ae) - (w_ce * P_ac * w_ae)
}
```

# Parameter values

The fixed parameters are the values used in the experiments.

```{r parameters}
w_ae <- 0.9
w_ce <- 0.3
b_a  <- 0.2

w_ac_vals <- seq(0, 1, by = 0.1)             # one line per value of w_ac
b_c_vals  <- seq(0, 1, length.out = 200)     # x-axis

df <- expand.grid(b_c = b_c_vals, w_ac = w_ac_vals)
df$P_ec_gen  <- with(df, P_ec_cccf(b_c, w_ac, w_ae, w_ce, b_a))
df$P_ec_prev <- with(df, P_ec_cc_minus_cf(b_c, w_ac, w_ae, w_ce, b_a))

# Annotation of the fixed parameters
param_label <- bquote(
  w[ae] == .(w_ae) * "," ~
  w[ce] == .(w_ce) * "," ~
  b[a]  == .(b_a)
)
```

# Figure A1

```{r figure-A1, fig.width = 12, fig.height = 6}
plot_pec <- function(yvar, title) {
  ggplot(df, aes(x = b_c, y = .data[[yvar]], color = factor(w_ac))) +
    geom_line(linewidth = 1.2) +
    labs(
      x = expression(b[c]),
      y = expression(P(e~"|"~c)),
      color = expression(w[ac]),
      title = title
    ) +
    annotate(
      "text",
      x = 0.02, y = 0.98,
      label = as.expression(param_label),
      hjust = 0, vjust = 1,
      size = 5
    ) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
    theme_minimal(base_size = 18) +
    theme(plot.title = element_text(size = 18))
}

gen  <- plot_pec("P_ec_gen",  "Generative Confounding")
prev <- plot_pec("P_ec_prev", "Preventive Confounding")

figureA1 <- ggarrange(gen, prev,
                      labels = c("A", "B"),
                      font.label = list(size = 24, color = "black"),
                      ncol = 2, nrow = 1)
figureA1

ggsave("figures/FigureA1.pdf", figureA1, width = 12, height = 6)
```

# Session info

```{r session-info}
sessionInfo()
```
