---
title: "Experiment 2: Analysis"
subtitle: "Making Predictions When Causal Models Change – The Role of Confounding"
format:
  html:
    embed-resources: true
    toc: true
    fig-dpi: 300
execute:
  warning: false
  message: false
---

This script reproduces all results reported for Experiment 2 (Section 4.2 and
the Participants section 4.1.1) and Figure 8 of the manuscript. Data:
`data/exp2_data.csv` (see `CODEBOOK.md`). Figure output: `figures/`.

# Setup

```{r setup}
library(tidyverse)  # data wrangling (dplyr, tidyr) and ggplot2
library(afex)       # aov_car()
library(emmeans)    # estimated marginal means and contrasts
library(binom)      # Wilson CIs for the cluster proportions
library(Hmisc)      # mean_cl_boot() used in stat_summary()
library(ggpubr)     # ggarrange()
library(showtext)   # font rendering

# Fonts: Google fonts are downloaded at run time. Without an internet
# connection the figure falls back to the default sans font.
fonts_ok <- tryCatch({
  font_add_google("Outfit", "title_font")
  font_add_google("Cabin", "body_font")
  TRUE
}, error = function(e) FALSE)
title_font <- if (fonts_ok) "title_font" else "sans"
body_font  <- if (fonts_ok) "body_font"  else "sans"
showtext_auto()
showtext_opts(dpi = 300)

dir.create("figures", showWarnings = FALSE)
```

# Data and exclusions

As pre-registered, participants who needed more than three rounds of the
comprehension test (`instr_tests > 3`) were excluded. The device and attention
confirmations are applied as in Experiments 1 and 3 (all participants
confirmed both).

```{r data}
tdata_raw <- read.csv("data/exp2_data.csv")

# rounds of the comprehension test needed, by condition
table(tdata_raw$causation_type, tdata_raw$instr_tests)

tdata <- subset(tdata_raw,
                desktop_conf == "1: I confirm" &
                attent_conf  == "1: I will take it seriously" &
                instr_tests  <= 3)

cat("Recruited:", nrow(tdata_raw),
    "| Excluded:", nrow(tdata_raw) - nrow(tdata),
    "| Final N:", nrow(tdata), "\n")

# n per condition
table(tdata$causation_type)
```

## Demographics

```{r demographics}
tdata %>%
  summarise(M_age = mean(age), SD_age = sd(age),
            min_age = min(age), max_age = max(age))
```

# Normative values

Normative values of P(e|c) for the models in Figure 6: Equation 4
(generative confounding link) and Equation A3 (preventive confounding link).

```{r normative}
# Equation 4: generative confounding link A -> C
p_ec_cccf <- function(w_ce, w_ae, w_ac, b_a, b_c) {
  numerator   <- w_ac * b_a + b_c * b_a - w_ac * b_c * b_a
  denominator <- w_ac * b_a + b_c - w_ac * b_c * b_a
  w_ce + (1 - w_ce) * w_ae * (numerator / denominator)
}

# Equation A3: preventive confounding link A -> C
# (b_c cancels out of P(a|c), Equation A1)
p_ec_prev <- function(w_ce, w_ae, w_ac, b_a) {
  p_a_given_c <- (b_a * (1 - w_ac)) / (1 - b_a * w_ac)
  w_ce + (1 - w_ce) * w_ae * p_a_given_c
}

preds <- data.frame(
  causation_type = c("generative", "generative", "preventive", "preventive"),
  case           = c("wac = 0.1", "wac = 0.9", "wac = 0.1", "wac = 0.9"),
  prediction     = c(
    p_ec_cccf(w_ce = 0.3, w_ae = 0.9, w_ac = 0.1, b_a = 0.2, b_c = 0.2),
    p_ec_cccf(w_ce = 0.3, w_ae = 0.9, w_ac = 0.9, b_a = 0.2, b_c = 0.2),
    p_ec_prev(w_ce = 0.3, w_ae = 0.9, w_ac = 0.1, b_a = 0.2),
    p_ec_prev(w_ce = 0.3, w_ae = 0.9, w_ac = 0.9, b_a = 0.2)
  )
)
preds$causation_type <- factor(preds$causation_type,
                               levels = c("generative", "preventive"))
preds$case <- factor(preds$case, levels = c("wac = 0.1", "wac = 0.9"))
preds
```

# Mean ratings

`case_01` is the rating for the model with the weak confounding link
(w_ac = 0.1), `case_09` the rating for the model with the strong link
(w_ac = 0.9), irrespective of the side on which the model was shown.

```{r long-format}
tdata_main <- tdata %>%
  select(subj_code, causation_type, side_order, case_01, case_09) %>%
  pivot_longer(cols = c(case_01, case_09),
               names_to = "case", values_to = "rating")

tdata_main$causation_type <- factor(tdata_main$causation_type,
                                    levels = c("generative", "preventive"))
tdata_main$case <- factor(tdata_main$case,
                          levels = c("case_01", "case_09"),
                          labels = c("wac = 0.1", "wac = 0.9"))

tdata_main$rating <- tdata_main$rating * 0.01  # convert to 0-1 scale
```

2 (Condition; between) × 2 (Model; within) ANOVA. The η²p of the interaction
is the effect size reported for the interaction contrast.

```{r anova}
a1 <- aov_car(rating ~ causation_type * case + Error(subj_code / case),
              tdata_main, anova_table = list(es = "pes"))
a1
```

Cell means (shown in Figure 8A):

```{r means}
emm <- emmeans(a1, ~ case * causation_type)
emm
```

Planned directed interaction contrast: ratings increase from weak to strong
w_ac in the generative condition and decrease in the preventive condition.

```{r interaction-contrast}
# cell order: (0.1, generative), (0.9, generative), (0.1, preventive), (0.9, preventive)
contrast(emm, list(interaction = c(-1, 1, 1, -1)), side = ">")
```

Planned simple contrasts within each condition: ratings for w_ac = 0.9 minus
ratings for w_ac = 0.1 (two-sided, as pre-registered).

```{r simple-contrasts}
emm_by_type <- emmeans(a1, ~ case | causation_type)
pairs(emm_by_type, reverse = TRUE, adjust = "none")
```

# Clusters

Participants are clustered by their intra-individual rating difference
(`case_09 - case_01`). Differences within ±3 points on the 0–100 rating scale
are classified as "no change".

```{r clusters}
margin_of_error <- 3  # on the 0-100 rating scale

tdata_clusters <- tdata %>%
  mutate(
    rating_difference = case_09 - case_01,
    cluster = case_when(
      abs(rating_difference) <= margin_of_error ~ "no change",
      rating_difference >  margin_of_error      ~ "increase",
      rating_difference < -margin_of_error      ~ "decrease"
    )
  )

tdata_clusters$causation_type <- factor(tdata_clusters$causation_type,
                                        levels = c("generative", "preventive"))

cluster_summary <- tdata_clusters %>%
  count(causation_type, cluster, name = "n") %>%
  group_by(causation_type) %>%
  mutate(n_total = sum(n),
         p = n / n_total,
         percentage = 100 * p) %>%
  ungroup() %>%
  # Wilson 95% CI for each bar
  bind_cols(
    binom.confint(x = .$n, n = .$n_total, methods = "wilson") %>%
      transmute(ci_low = 100 * lower, ci_high = 100 * upper)
  )

cluster_summary
```

## Cluster tests

Distribution of clusters across conditions (Pearson's χ² test):

```{r cluster-chisq}
cluster_table <- table(tdata_clusters$causation_type,
                       factor(tdata_clusters$cluster,
                              levels = c("decrease", "increase", "no change")))
cluster_table

chisq.test(cluster_table)
```

Distribution of clusters within each condition (χ² goodness-of-fit tests
against equal cluster frequencies):

```{r cluster-chisq-within}
chisq.test(cluster_table["generative", ])
chisq.test(cluster_table["preventive", ])
```

Among participants who changed their ratings, is the normative cluster larger
than the non-normative one? Directed (one-sided) binomial tests against 0.5,
restricted to the "increase" and "decrease" clusters.

```{r cluster-binom-direction}
# generative: "increase" > "decrease"
binom.test(x = cluster_table["generative", "increase"],
           n = cluster_table["generative", "increase"] +
               cluster_table["generative", "decrease"],
           p = 0.5, alternative = "greater")

# preventive: "decrease" > "increase"
binom.test(x = cluster_table["preventive", "decrease"],
           n = cluster_table["preventive", "decrease"] +
               cluster_table["preventive", "increase"],
           p = 0.5, alternative = "greater")
```

Normative cluster vs. "no change" cluster. Two-sided binomial tests against
0.5, restricted to these two clusters.

```{r cluster-binom-nochange}
# generative: "increase" vs. "no change"
binom.test(x = cluster_table["generative", "increase"],
           n = cluster_table["generative", "increase"] +
               cluster_table["generative", "no change"],
           p = 0.5)

# preventive: "decrease" vs. "no change"
binom.test(x = cluster_table["preventive", "decrease"],
           n = cluster_table["preventive", "decrease"] +
               cluster_table["preventive", "no change"],
           p = 0.5)
```

# Figure 8

```{r figure-8, fig.width = 6.9, fig.height = 8}
set.seed(1)  # bootstrapped CIs and jittered individual lines

pub_palette <- c("#E69F00", "#56B4E9", "#009E73",
                 "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# --- ONE knob: all text scales from BASE (identical to Exp 1) ----
BASE <- 14
LAB  <- BASE + 4    # panel letters A / B
TXT  <- BASE - 1    # in-plot percentage labels

theme_pub <- function(base = BASE) {
  theme_minimal(base_size = base) +
    theme(
      text             = element_text(family = body_font),
      strip.background = element_rect(fill = "grey92", color = "grey92"),
      strip.text       = element_text(face = "bold", size = base,
                                      family = title_font,
                                      margin = margin(5, 5, 5, 5)),
      axis.title       = element_text(face = "bold", size = base + 1),
      axis.title.x     = element_text(margin = margin(t = 10), face = "bold"),
      axis.title.y     = element_text(margin = margin(r = 10), face = "bold"),
      axis.text        = element_text(size = base, color = "grey20"),
      panel.grid.minor = element_blank(),
      legend.position  = "none"
    )
}

# --- Panel A: ratings --------------------------------------------
g3 <- ggplot(tdata_main, aes(x = case, y = rating,
                             fill = causation_type, color = causation_type)) +
  facet_wrap(~ causation_type, nrow = 1) +
  geom_line(aes(group = subj_code),
            position = position_jitter(height = 0.02, width = 0.05),
            linewidth = 0.35, alpha = 0.2) +
  geom_violin(alpha = 0.2, color = NA, trim = TRUE) +
  stat_summary(aes(group = 1), fun.data = mean_cl_boot, geom = "errorbar",
               width = 0, linewidth = 1) +
  stat_summary(aes(group = 1), fun = mean, geom = "line", linewidth = 1.3) +
  stat_summary(aes(group = 1), fun = mean, geom = "point", shape = 21,
               size = 4.2, stroke = 1.4, fill = "white") +
  geom_point(data = preds, aes(x = case, y = prediction),
             color = "black", fill = "black", size = 3, shape = 22) +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.25),
                     expand = c(0, 0)) +
  scale_color_manual(values = pub_palette) +
  scale_fill_manual(values = pub_palette) +
  labs(x = "Parameter Value", y = "P(e|c) Rating") +
  theme_pub() +
  theme(panel.spacing = unit(1.4, "lines"))

# --- Panel B: clusters -------------------------------------------
g4 <- ggplot(cluster_summary, aes(x = cluster, y = percentage,
                                  fill = causation_type)) +
  facet_wrap(~ causation_type, nrow = 1) +
  geom_bar(stat = "identity", alpha = 0.85, color = NA) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2,
                linewidth = 0.8, color = "grey20") +
  geom_text(aes(label = paste0(round(percentage, 0), "%")),
            vjust = -3.5, family = body_font, size = TXT / .pt) +
  scale_y_continuous(limits = c(0, 122), breaks = seq(0, 100, 25),
                     expand = c(0, 0)) +
  scale_fill_manual(values = pub_palette) +
  labs(x = "Cluster", y = "Percentage (%)") +
  theme_pub() +
  theme(panel.spacing = unit(1.4, "lines"))

# --- Assemble ----------------------------------------------------
figure8 <- ggarrange(g3, g4,
                     labels = c("A", "B"),
                     font.label = list(size = LAB, color = "black",
                                       family = title_font),
                     ncol = 1, nrow = 2, align = "v")
figure8

ggsave("figures/Figure8_Exp2.pdf", figure8,
       width = 6.9, height = 8, units = "in", device = cairo_pdf)
```

# Session info

```{r session-info}
sessionInfo()
```
