---
title: "Experiment 3: Analysis"
subtitle: "Making Predictions When Causal Models Change – The Role of Confounding"
format:
  html:
    embed-resources: true
    toc: true
    fig-dpi: 300
execute:
  warning: false
  message: false
---

This script reproduces all results reported for Experiment 3 (Section 5.2 and
the Participants section 5.1.1) and Figure 10 of the manuscript. Data:
`data/exp3_data.csv` (see `CODEBOOK.md`). Figure output: `figures/`.

# Setup

```{r setup}
library(tidyverse)  # data wrangling (dplyr, tidyr) and ggplot2
library(lme4)       # linear mixed-effects models
library(lmerTest)   # lmer() with F-tests (Satterthwaite df) for anova()
library(emmeans)    # planned contrasts
library(pbkrtest)   # Kenward-Roger degrees of freedom used by emmeans
library(binom)      # Wilson CIs for the cluster proportions
library(Hmisc)      # mean_cl_boot() used in stat_summary()
library(ggpubr)     # ggarrange()
library(showtext)   # font rendering

# Fonts: Google fonts are downloaded at run time. Without an internet
# connection the figure falls back to the default sans font.
fonts_ok <- tryCatch({
  font_add_google("Outfit", "title_font")
  font_add_google("Cabin", "body_font")
  TRUE
}, error = function(e) FALSE)
title_font <- if (fonts_ok) "title_font" else "sans"
body_font  <- if (fonts_ok) "body_font"  else "sans"
showtext_auto()
showtext_opts(dpi = 300)

dir.create("figures", showWarnings = FALSE)
```

# Data and exclusions

As pre-registered, participants who needed more than three rounds of the
comprehension questions were excluded. Each round consisted of two questions
(`ba_check`, `bc_check`), and `instr_tests` is incremented once per answered
question, so more than three rounds corresponds to `instr_tests > 6`.

```{r data}
tdata_raw <- read.csv("data/exp3_data.csv")

tdata <- subset(tdata_raw,
                desktop_conf == "1: I confirm" &
                attent_conf  == "1: I will take it seriously" &
                instr_tests  <= 6)

cat("Recruited:", nrow(tdata_raw),
    "| Excluded:", nrow(tdata_raw) - nrow(tdata),
    "| Final N:", nrow(tdata), "\n")

# n per condition
table(tdata$structure, tdata$parameter)
```

## Demographics

```{r demographics}
tdata %>%
  summarise(M_age = mean(age), SD_age = sd(age),
            min_age = min(age), max_age = max(age))
```

# Normative values

Normative values of P(e|c) for the models in Figure 9: Equation 2
(independent-causes model) and Equation 4 (confounded model).

```{r normative}
# Equation 2: independent-causes model
p_ec_ic <- function(w_ce, w_ae, b_a) {
  w_ce + b_a * w_ae - w_ce * b_a * w_ae
}

# Equation 4: confounded model (generative confounding link A -> C)
p_ec_cccf <- function(w_ce, w_ae, w_ac, b_a, b_c) {
  numerator   <- w_ac * b_a + b_c * b_a - w_ac * b_c * b_a
  denominator <- w_ac * b_a + b_c - w_ac * b_c * b_a
  w_ce + (1 - w_ce) * w_ae * (numerator / denominator)
}

preds <- data.frame(
  structure = rep(c("icce", "cccf"), each = 4),
  parameter = rep(c("bc", "bc", "ba", "ba"), times = 2),
  case      = rep(c("0.2", "0.9"), times = 4),
  prediction = c(
    # independent-causes model (b_c does not enter Equation 2)
    p_ec_ic(w_ce = 0.3, w_ae = 0.9, b_a = 0.2),                         # b_c = 0.2
    p_ec_ic(w_ce = 0.3, w_ae = 0.9, b_a = 0.2),                         # b_c = 0.9
    p_ec_ic(w_ce = 0.3, w_ae = 0.9, b_a = 0.2),                         # b_a = 0.2
    p_ec_ic(w_ce = 0.3, w_ae = 0.9, b_a = 0.9),                         # b_a = 0.9
    # confounded model
    p_ec_cccf(w_ce = 0.3, w_ae = 0.9, w_ac = 0.9, b_a = 0.2, b_c = 0.2), # b_c = 0.2
    p_ec_cccf(w_ce = 0.3, w_ae = 0.9, w_ac = 0.9, b_a = 0.2, b_c = 0.9), # b_c = 0.9
    p_ec_cccf(w_ce = 0.3, w_ae = 0.9, w_ac = 0.9, b_a = 0.2, b_c = 0.2), # b_a = 0.2
    p_ec_cccf(w_ce = 0.3, w_ae = 0.9, w_ac = 0.9, b_a = 0.9, b_c = 0.2)  # b_a = 0.9
  )
)
preds$structure <- factor(preds$structure, levels = c("icce", "cccf"))
preds$parameter <- factor(preds$parameter, levels = c("bc", "ba"))
preds$case      <- factor(preds$case, levels = c("0.2", "0.9"))
preds
```

# Mean ratings

`case_02` is the rating for the model in which the manipulated base rate
equals 0.2, `case_09` the rating for the model in which it equals 0.9,
irrespective of the side on which the model was shown.

```{r long-format}
tdata_main <- tdata %>%
  select(subj_code, structure, parameter, case_02, case_09) %>%
  pivot_longer(cols = c(case_02, case_09),
               names_to = "case", values_to = "rating")

tdata_main$structure <- factor(tdata_main$structure,
                               levels = c("ic", "cccf"),
                               labels = c("icce", "cccf"))
tdata_main$parameter <- factor(tdata_main$parameter, levels = c("bc", "ba"))
tdata_main$case <- factor(tdata_main$case,
                          levels = c("case_02", "case_09"),
                          labels = c("0.2", "0.9"))

tdata_main$rating <- tdata_main$rating * 0.01  # convert to 0-1 scale
```

Linear mixed-effects model with Causal Model, Changing base rate parameter,
and Base rate value as fixed effects and random intercepts for participants.

```{r lmm}
model <- lmer(rating ~ structure * parameter * case + (1 | subj_code),
              data = tdata_main)
```

Omnibus F-tests (Type III, Satterthwaite degrees of freedom):

```{r lmm-anova}
anova(model)
```

Planned contrasts (Kenward-Roger degrees of freedom).

b_a conditions: ratings for b_a = 0.2 minus ratings for b_a = 0.9, separately
for each causal model.

```{r contrasts-ba}
em_ba <- emmeans(model, ~ case | structure, at = list(parameter = "ba"),
                 lmer.df = "kenward-roger")
summary(pairs(em_ba), infer = TRUE)
```

b_c conditions: ratings for b_c = 0.9 minus ratings for b_c = 0.2, separately
for each causal model.

```{r contrasts-bc}
em_bc <- emmeans(model, ~ case | structure, at = list(parameter = "bc"),
                 lmer.df = "kenward-roger")
summary(pairs(em_bc, reverse = TRUE), infer = TRUE)
```

# Clusters

Participants are clustered by their intra-individual rating difference
(`case_09 - case_02`). Differences within ±3 points on the 0–100 rating scale
are classified as "no change".

```{r clusters}
margin_of_error <- 3  # on the 0-100 rating scale

tdata_clusters <- tdata %>%
  mutate(
    rating_difference = case_09 - case_02,
    cluster = case_when(
      abs(rating_difference) <= margin_of_error ~ "no change",
      rating_difference >  margin_of_error      ~ "increase",
      rating_difference < -margin_of_error      ~ "decrease"
    )
  )

tdata_clusters$structure <- factor(tdata_clusters$structure,
                                   levels = c("ic", "cccf"),
                                   labels = c("icce", "cccf"))
tdata_clusters$parameter <- factor(tdata_clusters$parameter,
                                   levels = c("bc", "ba"))

cluster_summary <- tdata_clusters %>%
  count(structure, parameter, cluster, name = "n") %>%
  group_by(structure, parameter) %>%
  mutate(n_total = sum(n),
         p = n / n_total,
         percentage = 100 * p) %>%
  ungroup() %>%
  # Wilson 95% CI for each bar
  bind_cols(
    binom.confint(x = .$n, n = .$n_total, methods = "wilson") %>%
      transmute(ci_low = 100 * lower, ci_high = 100 * upper)
  )

cluster_summary
```

# Figure 10

```{r figure-10, fig.width = 12, fig.height = 10}
set.seed(1)  # bootstrapped CIs and jittered individual lines

pub_palette <- c("#E69F00", "#56B4E9", "#009E73",
                 "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# --- ONE knob: all text scales from BASE (identical to Exp 1 and 2)
BASE  <- 14
LAB   <- BASE + 8   # panel letters A / B
TXT   <- BASE + 2   # in-plot percentage labels
STRIP <- BASE + 2   # facet strips (two-line)

theme_pub <- function(base = BASE) {
  theme_minimal(base_size = base) +
    theme(
      text             = element_text(family = body_font),
      strip.background = element_rect(fill = "grey92", color = "grey92"),
      strip.text       = element_text(face = "bold", size = STRIP,
                                      family = title_font,
                                      lineheight = 0.95,
                                      margin = margin(5, 4, 5, 4)),
      axis.title       = element_text(face = "bold", size = base + 1),
      axis.title.x     = element_text(margin = margin(t = 10), face = "bold"),
      axis.title.y     = element_text(margin = margin(r = 10), face = "bold"),
      axis.text        = element_text(size = base, color = "grey20"),
      panel.grid.minor = element_blank(),
      legend.position  = "none"
    )
}

# --- Facet labels ------------------------------------------------
struct_labels <- c(
  "icce" = "Independent-causes\nmodel",
  "cccf" = "Confounded\nmodel"
)

facet_lab <- labeller(
  structure = struct_labels,
  parameter = function(x) paste0("Parameter: ", x)
)

# --- Panel A: ratings --------------------------------------------
g3 <- ggplot(tdata_main, aes(x = case, y = rating,
                             fill = structure, color = structure)) +
  facet_wrap(~ parameter + structure, labeller = facet_lab, nrow = 1) +
  geom_line(aes(group = subj_code),
            position = position_jitter(height = 0.02, width = 0.05),
            linewidth = 0.35, alpha = 0.2) +
  geom_violin(alpha = 0.2, color = NA, trim = TRUE) +
  stat_summary(aes(group = 1), fun.data = mean_cl_boot, geom = "errorbar",
               width = 0, linewidth = 1) +
  stat_summary(aes(group = 1), fun = mean, geom = "line", linewidth = 1.3) +
  stat_summary(aes(group = 1), fun = mean, geom = "point", shape = 21,
               size = 4.2, stroke = 1.4, fill = "white") +
  geom_point(data = preds, aes(x = case, y = prediction),
             color = "black", fill = "black", size = 3, shape = 22) +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.25),
                     expand = c(0, 0)) +
  scale_color_manual(values = pub_palette) +
  scale_fill_manual(values = pub_palette) +
  labs(x = "Parameter Value", y = "P(e|c) Rating") +
  theme_pub() +
  theme(panel.spacing = unit(1.2, "lines"))

# --- Panel B: clusters -------------------------------------------
g4 <- ggplot(cluster_summary, aes(x = cluster, y = percentage,
                                  fill = structure)) +
  facet_wrap(~ parameter + structure, labeller = facet_lab, nrow = 1) +
  geom_bar(stat = "identity", alpha = 0.85, color = NA) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2,
                linewidth = 0.8, color = "grey20") +
  geom_text(aes(label = paste0(round(percentage, 0), "%")),
            vjust = -3, family = body_font, size = TXT / .pt) +
  scale_y_continuous(limits = c(0, 122), breaks = seq(0, 100, 25),
                     expand = c(0, 0)) +
  scale_fill_manual(values = pub_palette) +
  labs(x = "Cluster", y = "Percentage (%)") +
  theme_pub() +
  theme(panel.spacing = unit(1.2, "lines"))

# --- Assemble ----------------------------------------------------
figure10 <- ggarrange(g3, g4,
                      labels = c("A", "B"),
                      font.label = list(size = LAB, color = "black",
                                        family = title_font),
                      ncol = 1, nrow = 2, align = "v")
figure10

ggsave("figures/Figure10_Exp3.pdf", figure10,
       width = 12, height = 10, units = "in", device = cairo_pdf)
```

# Session info

```{r session-info}
sessionInfo()
```
